---
title: "Week 2 Tutorial: Normal distribution, t-distribution and hypothesis testing -- Solutions"
subtitle: Introduction to Statistical Thinking and Data Analysis
author: MSc in Epidemiology and MSc in Health Data Analytics, Imperial College London
date: 14 October 2019; reviewed 21 October 2019
output:
  pdf_document: 
    keep_tex: yes
header-includes:
- \usepackage{xcolor}
- \usepackage[utf8]{inputenc}
- \definecolor{darkred}{RGB}{139,0,0}
---

```{r setup, echo = FALSE}
my_plot_hook <- function(x, options)
  paste("\n", knitr::hook_plot_tex(x, options), "\n")
knitr::knit_hooks$set(plot = my_plot_hook)
```

1. Revisit the `perulung_ems.csv` dataset from the week 1 tutorial sheet, introduced on page 27 of Kirkwood and Sterne.
    
    | Variable     | Description                                                      |
    |--------------| -----------------------------------------------------------------|
    | id           | Participant ID number                                            |
    | fev1         | Forced Expiratory Volume in 1 second                             |
    | age          | Age in years                                                     |
    | height       | Height in centimeters                                            |
    | sex          | Sex (0 = female, 1 = male)                                       |
    | respsymptoms | Presence of respiratory symptoms (0 = no symptoms; 1 = symptoms) |
    
    a. Calculate an estimate and 95% confidence interval for each of the following quantities:

       \color{darkred}
       ```{r, message = FALSE}
       library(tidyverse)
       perulung <- read.csv("perulung_ems.csv")
       perulung <- perulung %>%
         mutate(sex = factor(sex, c(0, 1), c("female", "male")),
                respsymptoms = factor(respsymptoms, c(0, 1), c("no symptoms", "symptoms")))
       ```
       \color{black}

       i. Height in the whole population.
       ii. FEV1 in the population
       iii. Height for male and females children separately.
       iv. FEV1 for children with respiratory symptoms and those without respiratory symptoms.
       
        For each of these outcomes calculate the large sample CI using the normal distribution and the small-sample CI using the t-distribution. Compare your calculations to the 95% CI estimate output by the `t.test()` function.

       \color{darkred}
       ```{r, results = "hold"}
       x <- perulung$height

       ## A function to return requested outputs as a vector
       calc_ex1a <- function(x, digits = 3) {
         mu <- mean(x)
         se <- sd(x) / sqrt(length(x))
       
         print_ci <- function(ci) paste0("(", round(ci[1], digits), ", ", 
                                         round(ci[2], digits), ")")
        
         c(mean = round(mu, digits),  
           ci_norm = print_ci(mu + c(-1, 1) * qnorm(0.975) * se),
           ci_tdist = print_ci(mu + c(-1, 1) * qt(0.975, length(x)-1) * se),
           ci_t.test = print_ci(t.test(x)$conf.int))
       }
       
       rbind(
         height = calc_ex1a(perulung$height),
         fev1 = calc_ex1a(perulung$fev1),
         height_female  = calc_ex1a(perulung$height[perulung$sex == "female"]),
         height_male  = calc_ex1a(perulung$height[perulung$sex == "male"]),
         fev1_nosymp  = calc_ex1a(perulung$fev1[perulung$respsymptoms == "no symptoms"]),
         fev1_symptom  = calc_ex1a(perulung$fev1[perulung$respsymptoms == "symptoms"])
       ) %>%
       knitr::kable(align = "lcccc")
       ```
       
       The confidence intervals reported by `t.test()` are exactly the same as the small-sample 95% CI. They are very slightly wider than the large-sample CIs, but both are very close because the sample size is large.
       \color{black}

```{r, echo = FALSE}
print_ci <- function(ci, digits = 3) paste0("(", round(ci[1], digits), ", ", 
                                             round(ci[2], digits), ")")
```

    b. For each of the following research questions:
        
       * articulate an appropriate null and alternative hypothesis,
       * calculate an appropriate estimate and uncertainty range,
       * determine an appropriate statistical test for your hypothesis, and
       * report the results of your hypothesis test to answer the question.
    
       i. Is the average height of children aged 7 to 10 years in Lima greater than 124cm?
       
       \color{darkred}
       
        H0: The average height of children is $\leq124$cm.  
        H1: The average height of children is >124cm.
                
        We used a one-sample t-test with a one-sided alternative hypothesis that the average height is >124cm. We will reject the null hypothesis that if the _p_-value is less than $\alpha = 0.05$.
        
        The average height among children aged 7 to 10 was `r round(mean(perulung$height), 2)`cm with 95% confidence interval `r print_ci(t.test(perulung$height)$conf.int, 2)`, representating a range of values
        within which the true mean height likely falls. The _p_-value was `r round(t.test(perulung$height, mu = 124, alternative = "greater")$p.value, 3)`, indicating a reasonably high probability of observing a sample mean of `r round(mean(perulung$height), 3)` under the null hypthosis that height is $\leq 124$cm.  Thus we fail to reject the null hypothesis at the $\alpha = 0.05$ level.
         
        ```{r}
        mean(perulung$height)
        t.test(perulung$height)$conf.int
        t.test(perulung$height, mu = 124, alternative = "greater")
        ```
       \color{black}
       
       ii. Is the average height of girls in Lima equal to 123.5cm?
       
       \color{darkred}
       H0: Average height of girls in Lima is equal to\ 123.5cm.  
       H1: Average height of girls in Lima not equal to 123.5cm.  
       
       Test: One-sample t-test, two-sided alternative hypothesis.
       
       The average height among our sample of 335 girls aged 7 to 10 in Lima is 124.01cm with 95% confidence interval from 123.3cm to 124.7cm.  Compared to the reference value of 123.5cm, the t-statistic is 1.45 with 334 degrees of freedom, corresponding to a p-value of 0.147. Based on the 95% confidence range and/or the p-value, we conclude that our data are not inconsistent with the average height of girls aged 7 to 10 in Lima being 123.5cm
       
       _Note: we cannot conclude that the average height is 123.5cm (we can never prove the null hypothesis), only conclude that our data are not inconsistent with this value._
       
       ```{r}
       x_female <- perulung$height[perulung$sex == "female"]
       length(x_female)
       mean(x_female)
       t.test(x_female, mu = 123.5)
       
       ```
       \color{black}
       
       
       iii. Is there an association between sex and height amongst children in Lima?
       
       \color{darkred}
       
       H0: The average height of female children is equal to the average height of male children.
       H1: The average height of female children is not equal to the average height of male children.
       
       Hypothesis test: two-sample t-test with equal variance (unequal variance t-test also acceptable); two-sided alternative hypothesis.
       
       In our sample of 301 male children and 335 female children, the male children were 0.09cm taller than the female children (95% CI -0.89--1.06cm). The t-value was 0.172 on 634 degrees of freedom, corresponding to a two-sided p-value of 0.86. Is is very probable that we would have observed a difference this large or greater if there was no true height difference between male and female children. Thus we conclude that our data do not provide evidence to reject the null hypothesis that the average heights of male and female children are equal.
       
       ```{r}
       table(perulung$sex)
       x_female <- perulung$height[perulung$sex == "female"]
       x_male <- perulung$height[perulung$sex == "male"]
       
       mean(x_male) - mean(x_female)
       t.test(x_male, x_female, var.equal = TRUE)
       ```
       
       \color{black}
       
       iv. Is there an association between age and height amongst children in Lima?
       
       \color{darkred}
       We use linear regression to estimate the relationship between the outcome height as a linear function of age. The regression coefficient $\beta_1$ indicates the average change in height per additional year of age.
       
       H0: There is no association between height and age. $\beta_1 = 0$
       H1: There is an association between height and age. $\beta_1 != 0$
       
       Each additional year of age was associated with 5.2cm greater height, with 95% confidence interval from 4.6cm to 5.7cm.  The t-value was 18.6 on 634 degrees of freedom, with associated two-sided p-value less than 0.001.  This is strong evidence to reject the null hypothesis of no association between height and age, and conclude that height is positively associated with age amongst children aged 7 to 10.
       
       ```{r}
       fit <- lm(height ~ age, data = perulung)
       summary(fit)
       confint(fit)
       ```
       \color{black}
       
       v. Do children with respiratory symptoms have reduced pulmonary function compared to children with no respiratory symptoms?
       
       \color{darkred}
      
       H0: FEV1 for children with respitory symptoms is greater than or equal to FEV1 for children with no symptoms.  
       H1: FEV1 for chidren with respiratory symptoms is less than FEV1 for children with no symptoms.
       
       Hypothesis test: two-sample t-test, unequal variance (equal variance also justifiable), **one-sided** alternative hypothesis.
       
       The FEV1 for children with with respiratory symptoms was 0.15 litres/second lower than children with no respiratory symptoms (95% CI 0.09--0.21 litres/second). The t-statistic was -4.90 on 211.5 degrees of freedom. The corresponding one-sided _p_-value was <0.001, indicating strong evidence to reject the null hypothesis of no difference and conclude that FEV1 is statistically significantly lower for children with respiratory symptoms compared to those with no symptoms.
       
       ```{r}
       fev1_symp <- perulung$fev1[perulung$respsymptoms == "symptoms"]
       fev1_nosymp <- perulung$fev1[perulung$respsymptoms == "no symptoms"]
       mean(fev1_symp) - mean(fev1_nosymp)
       t.test(fev1_symp, fev1_nosymp)
       t.test(fev1_symp, fev1_nosymp, alternative = "less")
       ```
       \color{black}
       
       vi. Do children who are taller than average for their age also have greater pulmonary function?
       
       \color{darkred}
       We used multiple linear regression to estimate the relationship linear relationship between the outcome FEV1 and height, adjusted for age.
       
       H0: There is no association between height and FEV1, after adjusting for the effect of age on FEV1.
       H1: There is an independent effect of height on FEV1, after adjusting for the effect of age.
       
       Adjusted for age, each centimeter greater height was associated with 0.025 litres/second greater FEV1 (95% CI 0.021--0.029 liters/s; p <0.001). Thus our data provide strong evidence that greater height is associated with greater FEV1, even after adjusting for the positive association between of older age and FEV1.
       
       ```{r}
       fit6 <- lm(fev1 ~ height + age, data = perulung)
       summary(fit6)
       confint(fit6)
       
       ```
       
       \color{black}
    
    c. ~~What is the probability that your analyses have correctly answered all six of these questions?~~
       CORRECTED QUESTION: If the null hypothesis were true for all of the above six questions, what is the probability of erroneously rejecting at least one null hypothesis and incorrectly concluding an association exists?
       
       \color{darkred}
       Using a Type I error rate threshold of $\alpha = 0.05$, if the null hypothesis is true, there is a 95% probability of correctly failing to reject the null hypothesis and a 5% probability of erroneously rejecting the null hypothesis and incorrectly concluding that there is an association.
       
       If all six null hypotheses are true (and all of the tests are independent), the probability of correctly rejecting all of them is 0.95^6^ = 0.735.  Thus the probability that we erroneously reject at least one null hypothesis is 1 - 0.735 = 26.5%.
       \color{black}
\newpage

2. _Statistical modelling: using the normal distribution to estimate population distributions._ The purpose of this exercise is to practice applying the normal distribution to estimate the distribution of an outcome in a population using data from a sample drawn from the population. The version of the NHANES dataset we are using is constructed to approximate a simple random sample from the American population. Thus we can use the observations in our sample to make inference about the American population.

   We will use dataset `nhanes20pl` subsetted to the adult respondents aged 20 plus.

   ```{r}
   library(NHANES)
   data(NHANES)
   nhanes20pl <- NHANES[NHANES$Age >= 20, ]
   ```

   For each of the questions below, consider three continuous variables: 

   * `Height`: Standing height in centimeters.
   * `BMI`: Body Mass Index
   * `AlcoholYear`: Number of days over the past year that participant drank alcoholic beverages.

   a.  Create a graph of the frequency distribution for each of the variables, either using histograms or density plots. Does each outcome appear consistent with arising from a normal distribution?

      \color{darkred}
      Height appears normally distributed. BMI appears normally distributed perhaps with a slight positive skew. Alcohol days per year is severely positively skewed.
      ```{r, fig.height = 2.5, fig.width = 3, warning = FALSE}
      ggplot(nhanes20pl, aes(Height)) +
      geom_histogram(bins = 50)
      
      ggplot(nhanes20pl, aes(BMI)) +
      geom_histogram(bins = 50)
      
      ggplot(nhanes20pl, aes(AlcoholYear)) +
      geom_histogram(bins = 50)
      ``` 
      \color{black}
      
   b. Calculate and report the sample mean and standard deviation for each of the outcomes.
   
      \color{darkred}
      ```{r}
      height_mean <- mean(nhanes20pl$Height, na.rm = TRUE)
      height_sd <- sd(nhanes20pl$Height, na.rm = TRUE)
      bmi_mean <- mean(nhanes20pl$BMI, na.rm = TRUE)
      bmi_sd <- sd(nhanes20pl$BMI, na.rm = TRUE)
      alc_mean <- mean(nhanes20pl$AlcoholYear, na.rm = TRUE)
      alc_sd <- sd(nhanes20pl$AlcoholYear, na.rm = TRUE)
      
      data.frame(mean = c(height_mean, bmi_mean, alc_mean),
                 sd = c(height_sd, bmi_sd, alc_sd),
                 row.names = c("height", "bmi", "alcohol_year")) %>%
      knitr::kable(digits = 1, align = "cc")
      ```
      
      \color{black}
      
   
   c. Use the normal distribution as a model to estimate the following:

      i. The proportion of adults who are above 165cm tall.
      ii. The proportion of adults between 153cm and 160cm tall.
      iii. The height of a door in order that 90% of adults can walk under without ducking.
      iv. The proportion of adults who are obese, defined as BMI above 30.
      v. The proportion of adults who are overweight, defined as BMI between 25 and 30.
      vi. The BMI threshold at which 25% of adults are below.
      vii. The proportion adults who drink alcohol on more than 100 days per year.
      viii. The proportion of adults who drink alcohol on fewer than 10 days per year.
      ix. The interquartile range for the number of days per year that American adults drink alcohol.

      ```{r}
      ans2c <- c(1-pnorm(165, height_mean, height_sd),
                  pnorm(160, height_mean, height_sd) - pnorm(153, height_mean, height_sd),
                  qnorm(0.9, height_mean, height_sd),
                  1-pnorm(30, bmi_mean, bmi_sd),
                  pnorm(30, bmi_mean, bmi_sd) - pnorm(25, bmi_mean, bmi_sd),
                  qnorm(0.25, bmi_mean, bmi_sd),
                  1-pnorm(100, alc_mean, alc_sd),
                  pnorm(10, alc_mean, alc_sd),
                  qnorm(0.75, alc_mean, alc_sd) - qnorm(0.25, alc_mean, alc_sd))
      ```
      
   d. Check your estimates based on the normal distribution by directly calculating each of the above proportions amongst the observed sample. Based on this comparison, do you think that the normal distribution is a good model for the population distribution of each outcome in the population?

      \color{darkred}
      ```{r}
      ans2d <- c(mean(nhanes20pl$Height > 165, na.rm=TRUE),
                 mean(nhanes20pl$Height > 153 & nhanes20pl$Height < 160, na.rm=TRUE),
                 quantile(nhanes20pl$Height, 0.9, na.rm=TRUE),
                 mean(nhanes20pl$BMI > 30, na.rm=TRUE),
                 mean(nhanes20pl$BMI > 25 & nhanes20pl$BMI <= 30, na.rm=TRUE),
                 quantile(nhanes20pl$BMI, 0.25, na.rm=TRUE),
                 mean(nhanes20pl$AlcoholYear > 100, na.rm=TRUE),
                 mean(nhanes20pl$AlcoholYear <= 10, na.rm=TRUE),
                 diff(quantile(nhanes20pl$AlcoholYear, c(0.25, 0.75), na.rm=TRUE)))
          
      labels <- c("Height above 165cm", "Height between 153-160cm", "Height 90th percentile",
                   "BMI above 30", "BMI between 25-30", "BMI 25th percentile",
                   "Alcohol days above 100", "Alcohol days below 10", "Alcohol days IQR")
      data.frame(normal_approx = ans2c,
                 empirical_dist = ans2d,
                 row.names = labels) %>%
                 knitr::kable(digits = 2)
      ```

      \color{black}

\newpage

3. _Distribution of the sample mean and confidence intervals._ This exercise will use the NHANES dataset as 'true' population from which to simulate samples and study the properties of the sample mean as an estimator of the population mean. We will use the same three variables studied in exercise 2.  Start by creating a vector of responses removing any `NA` values.

   ```{r}
   height <- nhanes20pl$Height[!is.na(nhanes20pl$Height)]
   bmi <- nhanes20pl$BMI[!is.na(nhanes20pl$BMI)]
   alcohol <- nhanes20pl$AlcoholYear[!is.na(nhanes20pl$AlcoholYear)]
  ```

   Do the following steps for each variable: 
   
   a. Simulate a large number of samples with replacement of size 5, 10, 25, 50, 100, and 500. See the function `?sample`. (The function `?replicate` might also be helpful for simulating a large number of samples.)
   
   b. For each sample, calculate the sample mean, the large-sample 95% CI using the normal distribution, and the small-sample 95% CI using the t-distribution. (The function `?apply` might be useful.)
   
   c. Plot the distribution of the sample mean for each of the increasing sample size. What do you notice about the distribution of the sample mean with increasing sample size for each variable?
   
      \color{darkred}
      The distribution of the sample mean becomes more concentrated around the true population value as the sample size increases.  
      
      For height, which follows a normal distribution, the sample mean is also normally distributed for all sample sizes.
      
      For BMI, which has a slight positive skew, the sample mean also has a slight positive skew but becomes normally distributed for n = 25.
      
      For alcohol days per year, which is severely positively skewed, the sample mean is also very skewed at small sample sizes n=5 and n=10.
      \color{black}
   
   d. Calculate the coverage of the 95% confidence intervals by calculating for what proportion of samples the 95% confidence interval contains the true population mean (calculated in exercise 2b). Create a figure to visually summarise the analysis of 95% CI coverage at different sample sizes, for the large-sample and small-sample CI, and for different variables.
     
      i. How do the coverage of the large-sample CI (using normal distribution) and small-sample CI (using t-distribution) compare at different sample sizes?
      ii. For a given sample size, how does the coverage compare for each of the three variables? Can you relate this to your findings in exercise 2?

      \color{darkred}

      For height, both the normal and t-distribution based confidence intervals attain the nominal 95% coverage when n>50.  The calibration of the t-distribution confidence intervals is perfect (equal to 95%) for all sample sizes.
      
      For BMI, which is only slightly non-normally distributed, both attain 95% coverage for n>100 (and close for n>50). At sample sizes n $\leq$ 10, the small sample confidence interval is below the nominal 95% level, but very close at 93.6%.
      
      For alcohol consumption days per year, the CI coverage attains 95% for n = 500 and is above 93% for n $\geq$ 100. However, at small samples n $\leq$ 10, the coverage of the t-distribution based CI is better than the coverage of the normal distribution based CI, but is substantially below the nominal 95% level.  This is because the t-distribution approximation relies on assumption that the outcome is normally distributed in the population which is severely violated for the number of alcohol days per year outcome.
      
```{r, cache = TRUE}
set.seed(77316870)

nsim <- 10000 # number of simulated datasets
nsamp <- c(5, 10, 25, 50, 100, 500)   # sample size

sim_ex3 <- function(x, n) {

  mu <- mean(x)  # true population mean
  samp <- replicate(nsim, sample(x, n, replace = TRUE))

  est <- colMeans(samp)
  se <- apply(samp, 2, sd) / sqrt(n)

  ci_l_z <- est - qnorm(0.975) * se
  ci_u_z <- est + qnorm(0.975) * se

  ci_l_t <- est - qt(0.975, n-1) * se
  ci_u_t <- est + qt(0.975, n-1) * se
  
  list(est = est,
       cover_z = mean(ci_l_z < mu & mu < ci_u_z),
       cover_t = mean(ci_l_t < mu & mu < ci_u_t))
}

height_res <- list()
bmi_res <- list()
alc_res <- list()

for(n in nsamp){
  height_res[[as.character(n)]] <- sim_ex3(height, n)
  bmi_res[[as.character(n)]] <- sim_ex3(bmi, n)
  alc_res[[as.character(n)]] <- sim_ex3(alcohol, n)
}
```

Part (c):

```{r, fig.height = 3.5, fig.width=6, fig.show = "hold", cache = TRUE}
## sample mean summary
sample_mean_summary <-
  bind_rows(
    data.frame(outcome = "height", 
               sapply(height_res, "[[", "est"), 
               check.names = FALSE, stringsAsFactors = FALSE),
    data.frame(outcome = "bmi", 
               sapply(bmi_res, "[[", "est"), 
               check.names = FALSE, stringsAsFactors = FALSE),
    data.frame(outcome = "alcohol", 
               sapply(alc_res, "[[", "est"), 
               check.names = FALSE, stringsAsFactors = FALSE)
  ) %>%
  gather(n, sample_mean, `5`:`500`) %>%
  mutate(outcome = fct_relevel(outcome, "height", "bmi", "alcohol"),
         n = as.integer(n))
    
sample_mean_summary %>%
  filter(outcome == "height") %>%
  ggplot(aes(sample_mean)) +
  geom_histogram(aes(y = ..density..), bins = 100) +
  geom_vline(xintercept = mean(height), color = "red") +
  facet_wrap(~n, scales = "free_y") +
  theme_light() +
  ggtitle("Height: sample mean")

sample_mean_summary %>%
  filter(outcome == "bmi") %>%
  ggplot(aes(sample_mean)) +
  geom_histogram(aes(y = ..density..), bins = 100) +
  geom_vline(xintercept = mean(bmi), color = "red") +
  facet_wrap(~n, scales = "free_y") +
  theme_light() +
  ggtitle("BMI: sample mean")

sample_mean_summary %>%
  filter(outcome == "alcohol") %>%
  ggplot(aes(sample_mean)) +
  geom_histogram(aes(y = ..density..), bins = 100) +
  geom_vline(xintercept = mean(alcohol), color = "red") +
  facet_wrap(~n, scales = "free_y") +
  theme_light() +
  ggtitle("Alcohol days per year: sample mean")
```

Part (d):
```{r, fig.height = 2.5, fig.width=7, cache = TRUE}
## Compile coverage estimates
cover_summary <- 
  bind_rows(
    tibble(outcome = "height", 
           n = nsamp, 
           ci_type = "z", 
           coverage = sapply(height_res, "[[", "cover_z")),
    tibble(outcome = "bmi", 
           n = nsamp, 
           ci_type = "z",
           coverage = sapply(bmi_res, "[[", "cover_z")),
    tibble(outcome = "alcohol", 
           n = nsamp, 
           ci_type = "z",
           coverage = sapply(alc_res, "[[", "cover_z")),
    tibble(outcome = "height", 
           n = nsamp, 
           ci_type = "t", 
           coverage = sapply(height_res, "[[", "cover_t")),
    tibble(outcome = "bmi", 
           n = nsamp, 
           ci_type = "t",
           coverage = sapply(bmi_res, "[[", "cover_t")),
    tibble(outcome = "alcohol", 
           n = nsamp, 
           ci_type = "t",
           coverage = sapply(alc_res, "[[", "cover_t")),
    
  ) %>%
  mutate(outcome = fct_relevel(outcome, "height", "bmi", "alcohol"))

## summary table
cover_summary %>% 
  unite(key, outcome, ci_type) %>%
  spread(key, coverage) %>%
  select(n, height_z, height_t, bmi_z, bmi_t, alcohol_z, alcohol_t) %>%
  knitr::kable(caption = "Coverage of theoretical 95% confidence intervals")

## summary figure
cover_summary %>%
  mutate(n = as.factor(n)) %>%
  ggplot(aes(n, coverage, color = ci_type, group = ci_type)) +
  geom_hline(yintercept = 0.95, linetype = "dashed") +
  geom_point() +
  geom_line() +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~outcome) +
  theme_light()

```

\color{black}